Triple bottom line or trilemma? Global tradeoffs between prosperity, inequality, and the environment

### 2024-02-26 revised_manuscript_240226.Rmd
revisions round 4

- No country in the world simultaneously achieved high prosperity, low inequality, and low environmental impact between 1995-2017.
- A sustainability “trilemma” as opposed a “triple bottom line” appears to exist for international development.
- There is an “inequality trap” that appears to be particularly acute in Latin America.
- Country trajectories suggest that reducing inequality can be achieved without compromising the biosphere.
- The three dimensions need be considered together to identify international development challenges.

@@

A key aim of sustainable development is the joint achievement of prosperity, equality, and environmental integrity: in other words, material living standards that are high, broadly-distributed, and low-impact. This has often been called the “triple bottom line”. But instead, what if there is a " trilemma" that inhibits the simultaneous achievement of these three goals? We analyzed international patterns and trends in the relationships between per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint from 1995 to 2017. Based on these widely-used measures of development, we found that no country achieved the triple bottom line. Over the study period, countries tended to decrease in income inequality before growing in prosperity and environmental impact. A cluster of countries, primarily from Latin America, appeared trapped in a regime of high inequality, while the instances of prosperous countries lowering inequality only coincided with increasing environmental impact. The existence of country clusters and widely-divergent national trajectories show that common pair-wise explanations such as Kuznets Curves do not fully capture important development dynamics, especially in a period of global environmental change. Instead, frameworks need to consider all three dimensions together, and especially the relationship between inequality and the environment. Additionally, the past development trajectories of countries, how they fit into the structure of the world economy, and the role of social-ecological feedbacks in strengthening clustering effects such as “inequality traps”. Practically, our findings suggest that to overcome the sustainability trilemma, many countries - especially those that are already prosperous - should prioritise social redistribution and environmental stewardship before further economic growth.

@@

Although global prosperity has increased in recent generations, progress has been uneven. Even today, a significant share of the world’s population lives under the international poverty line (a standard that the UN itself recognizes may grossly undercount the world's impoverished [@Alston_20220]). Right above this derisory threshold, hundreds of millions more people maintain a precarious subsistence and aspire for higher and more secure living standards [@TheWor:2018]. And while average incomes for certain developing countries have been converging with those of advanced economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened across the world [@Bourguignon:2007; @Milanovic:2016]. This has made  inequality an urgent issue for policymakers and the general public due to  anxieties about social and political instability [@Chancel:2020; @UN2020]. Environmental integrity forms the third, and perhaps most fundamental, dimension of the sustainability [@arrow1995economic]. There is now a scientific consensus that human activities have had severely deleterious impacts on the biosphere, the health of which determines long-run human wellbeing [@Díaz:2019; @folke2021our]. Indeed, humanity has already exceeded a number of the biophysical boundaries that delimit a "safe operating space" for humanity [@Rockström:2009; @RaworthKate:2017; @Rockstr_m_2023].

@@

We used these data to chart the movement of countries across the dimensions of prosperity, inequality, and environmental impact over time. We further investigated whether these movements trace out typologies of country trajectories within the three-dimensional space. To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time. We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to determine the relevant number of clusters (clustering is a statistical process whereby objects are partitioned and then grouped in such a way that the members of one cluster are more closely related to each other than they are to those of other clusters). We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by Charrad _et al._, [-@Charrad:2014tp] and Brook _et al._ [-@Brock:2008vz]. We used principal component analysis as an ordination method for visualization, and then vector-fitting to show how clusters are pulled apart in the principal component space (in order to make the three-dimensional distribution of countries intelligible). All data used are publicly available, and the code used for the analysis is available at: <link will be added upon publication, avoided here due to double-blind requirements>.

While we were able to assess 140 countries "statically" (i.e., ignoring temporal dynamics), an analysis of the missing values showed that analyzing changes over time in the trilemma space was only possible for 59 countries. The most incomplete dataset is that for inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values. Based on the quality of the three datasets, we selected the time-series from 1995-2017 for the 59 countries with fewer than 30\% of missing values and then imputed the missing values with linear interpolation. This reduced dataset enables the study of the temporal dynamics of countries in the trilemma space (Fig \ref{fig:trilemma}).

@@

Our analysis found that no country simultaneously achieved high prosperity, low inequality, and low environmental impact. This strongly suggests that globally, there is a sustainability trilemma. At the country level, there were widespread tradeoffs between reducing inequality and reducing environmental impact (Fig \ref{fig:trilemma}E - as the Gini coefficient for income decreases, ecological footprint increases) as well as between increasing prosperity and reducing environmental impact (Fig \ref{fig:trilemma}G - per-capita GNI and ecological footprint move in the same direction). Results show synergies between increasing prosperity and decreasing inequality (Fig \ref{fig:trilemma}G), although a number of high-income countries (e.g., Sweden and New Zealand) exhibited low levels of inequality in the past but then high variation and a trend towards greater inequality in more recent years.

While these global patterns provide insight into international variations in key dimensions of sustainability, they do not completely reveal the diversity of individual country trajectories, nor their typologies. To further investigate country trajectories, we calculated the displacement within the three-dimensional space for each pair-wise dimension, where $\gamma$ (gamma) is the angle in the inequality and prosperity plane while $\theta$ (theta) is the angle in the inequality and environmental impact plane. These two angles and the total distance "travelled" in the three-dimensional space help to define typologies in country trajectories.

@@

There are certain regions in the three-dimensional space where countries tended to dwell (Fig \ref{fig:cube_ensamble}). Multiple stable states can exist if feedbacks between development drivers are present. Statistically, this translates into multi-modal distributions and so-called "regimes". The prosperity-environment plane (Fig \ref{fig:cube_ensamble}A), and inequality-environment plane (Fig \ref{fig:cube_ensamble}C) have two pronounced clusters of countries (the movement of countries during the study period stayed within a relatively circumscribed space), while the inequality-prosperity plane (Fig \ref{fig:cube_ensamble}B) has one major cluster and one looser grouping. Countries also showed distinctive profiles in direction of movement (Fig \ref{fig:cube_ensamble} insets). Most countries had low displacement (short arrows), which means their trajectory over the study period did not take them far from where they started. A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995. For example, between 1995 and 2017, China's Gini coefficient for income and per-capita ecological footprint worsened while per-capita GNI grew significantly. And while inequality and environmental impact fell in Ukraine, prosperity only increased marginally. To better understand country typologies, we clustered summary statistics (i.e., mean and standard deviations) of the three indicators, as well as the direction and magnitude of country movement over time (Fig \ref{fig:pca}).

Combining the three planes, the trajectories of countries from 1995 to 2017 can be categorized into three general types (Fig \ref{fig:pca}A). First, there was a group composed of countries with relatively high prosperity and ecological footprint; their movements were relatively less erratic and had larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group). This cluster can be called "Regime A". Next, there was a group of countries, mainly from Latin America, that seemed to be "trapped" in a region of high inequality during the study period. They moved toward regions of higher income inequality with respect to the sample and with a larger standard deviation of $\theta$; that is to say, large variations along the income inequality and ecological footprint axes (Fig \ref{fig:pca}A, in green). Within the three-dimensional space, this cluster of countries with high inequality and relatively low prosperity and environmental impact can be called "Regime B". A third group of countries showed large displacements within the trilemma space but in no definite direction, and was characterized by large variations in prosperity and inequality (Fig \ref{fig:pca}A, in yellow).

If we ignore temporal dynamics and focus on the mean and standard deviation of the indicators, we are able to include more countries (N=140). This static analysis also yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between prosperity, inequality, and environmental impact as in the findings from the dynamic analysis. The first cluster (Fig \ref{fig:pca}B, in yellow) was composed primarily of advanced Western economies and had high prosperity, inequality, and environmental impact. The second cluster (Fig \ref{fig:pca}B, in green) was composed of countries from Africa, Latin America, and Asia - broadly, the "Global South". Here, countries all tended to have lower prosperity (e.g., low- or middle-income status), but among them exhibited large variations in inequality and environmental impact.

In both the dynamic and the static analysis, decreasing inequality and decreasing environmental impact appear to pull countries in opposite directions (Fig \ref{fig:pca}), providing evidence for the difficulty of simultaneously achieving both goals. In Fig \ref{fig:pca}A, the fitted vectors (arrows) show the direction in which variables explain the variance in the data and thus countries' differences. While per-capita ecological footprint and GNI point in similar directions (countries tended to increase both at the same time), the Gini for income points in the opposite direction. This suggests that countries that reduce inequality do so after increasing in environmental impact and prosperity. In the ordination performed in the larger static sample (Fig \ref{fig:pca}B), we find a similar pattern. In fact, the trajectories analysis found that countries that reduced inequality subsequently increased in environmental impact (Fig \ref{fig:cube_ensamble}A). Reducing inequality was often followed by rising prosperity (Fig \ref{fig:cube_ensamble}B), although some middle-income countries maintained relatively low levels of inequality (at levels similar to higher-income countries) while still maintaining lower levels of environmental impact (Fig \ref{fig:cube_ensamble}B, C).

@@

Understanding the patterns and trends in the relationships between prosperity, inequality, and environmental impact is an important topic for sustainability science and development research. In the existing literatures of economic, social, and environmental science, various arguments have been advanced about the relationships between the three goals. Most commonly, mechanisms have been hypothesized for pair-wise relationships, with Kuznets Curve-type relationships being the most widely analyzed. The (Environmental) Kuznets Curve posits that growing prosperity leads to worsening inequality and environmental impact, before reaching an inflection point after which the relationship is reversed. The evidence for this hypothesis has been contested with respect to both inequality [@Chancel:2020][@Pikett:2014] and the environment [@stern2004rise; @stern1996economic ]. For the latter, conclusions have often varied based on the indicator chosen to proxy environmental impact [@MALER201325]. Our analysis did not detect KC-type relationships between prosperity and either the Gini coefficient for income or ecological footprint Fig \ref{fig:S5}) (please see SI for more details). Instead, the findings strongly suggest that growing prosperity is almost always associated with worsening inequality and environmental impact.

What is clear from earlier empirical studies of pair-wise relationships of prosperity, inequality, and the environment is that whatever tradeoffs and synergies that could be identified were mediated by exogenous factors [@dasgupta2002confronting; @pothen:2019]. For instance, in his survey of relevant studies, Chancel [-@Chancel:2020] concludes that what empirical evidence has been found to support (E)KC-type relationships are actually artefacts particular to post-World War II datasets or transient historical trends with limited applicability to today’s inequality dynamics. It is only by considering the economic, social, and environmental dimensions together, and the different ways in which countries have evolved with respect to them, can we arrive at a more comprehensive understanding. In particular, while the prosperity-inequality and prosperity-environment relationships have been widely analyzed (with mixed conclusions), the inequality-environment dynamic has been less explored – and should also be considered with respect to the potentially mediating effects of the third dimension of prosperity. In our findings, this necessity is underscored by the existence of international clusters that moved in distinctively different directions.

Overall, the low frequency with which countries were able to meet even two goals – no countries in the case of the prosperity-environment relationship – in both the dynamic and static analysis is notable. This suggests that there is indeed a sustainability trilemma facing global development. In other words, despite its prominence as a definition of sustainability, our study indicates it is highly difficult, and potentially infeasible, for countries to reach the triple bottom line of prosperity, equality, and environmental integrity. However, the burden of overcoming the trilemma is not even across countries. Another important finding of the analyses is revealed through national trajectories in the three-dimensional space. Over the study period, some countries, primarily from Latin America, struggled to escape areas characterised by high inequality, which may represent an "inequality trap" subtype of the trilemma.

The existence of the trilemma in general, and of the inequality trap in particular, could be better explained through an understanding of how inequality and environmental impact are related. A potentially useful approach is to investigate how different levels of income distribution impact environmental integrity at different levels of average income (thereby incorporating the prosperity dimension). In the existing literature on inequality-environment dynamics, two broad types of mechanisms have been posited: those that are "economic" (based on assumptions about the consumption decisions of rational individuals with different relative incomes) and those that are "political" (based on assumptions about how income distribution influences collective decision-making in the form of public policy) [@berthe2015mechanisms; @@Hamann:2018].

The "economic" (often called the "marginal propensity to emit") hypothesis argues that higher inequality lessens environmental impact because a more equal distribution raises overall consumption levels. This is based on the conventional microeconomic assumption that there is a higher marginal propensity to consume among the poor than among the rich. The political – sometimes "political economy" – hypothesis was first posited by Boyce [-@boyce1994inequality] and argues that by concentrating power in the hands of the rich, economic inequality worsens environmental conditions. One mechanism is that the well-heeled tend to have a vested interest in stopping regulations and laws that protect the environment (and thereby lower profits). This has been supported by agent-based modelling studies such as that by Kempf and Rossignol [-@KEMPF_2007]. Another hypothesized mechanism comes from the opposite side of the social spectrum, in which high inequality drives the poor to engage in more environmentally-destructive activities to meet their basic needs [@zhao2021analysis].

Some studies have found greater empirical evidence for the economic explanation than the political (economy) mechanisms. In a study of Sub-Saharan countries from 1995 to 2018, Gimba _et al._ [-@gimba2023towards] found that rising income inequality reduces ecological footprint but exerts a significant drag on economic growth – with critical implications for poverty alleviation in a region with many of the world’s lowest-income countries. In a study of Western African states from 1984 to 2016, Langnel _et al._ [-@langnel2021income] similarly found that income inequality often lowers ecological footprint. The "marginal propensity to emit" hypothesis also found empirical support in Ali [-@ali2023income], which analyzed found a tradeoff relationship between the Gini for income and carbon emissions in 42 middle-income countries between 1990 and 2016. In high-income economies, Hailemariam _et al._ [-@hailemariam2020carbon]’s study of OECD countries found that increasing income inequality (also measured by Gini) was associated with decreasing carbon emissions.

However, there has also been strong empirical evidence for the opposite effect, with lower inequality corresponding to lower environmental impact. In a major early study, Holland _et al._ [-@holland2009cross] found that high income inequality (measured by Gini) was a significant factor predicting anthropogenic biodiversity loss. In a study of 135 countries over the period 1981-2012, Malerba [-@malerba2020trade] found that reducing income inequality consistently reduced the tradeoff between rising prosperity and environmental impact as measured by the "carbon intensity of poverty reduction." In a study of five middle-income ASEAN countries, Masud _et al._ [-@masud2020does] found that from 1985-2015, a tight relationship between income inequality and environmental degradation, and with a bidirectional causality for the bottom 40% of the income distribution. Overall, they concluded that "policymakers should reduce income inequality to strengthen environmental sustainability". Focusing on the relationships between carbon emissions, economic growth, and income inequality before and after the 2008 "Great Recession" across 70 countries, a "robust finding" by Agusdinata _et al._ [-@agusdinata2021economic] was that countries on trajectories of declining emissions were also on trajectories of declining inequality.

The case for the inequality-environment relationship being segmented by development level (i.e., average national income) has also found evidence in analyses of global datasets. Ehigiamusoe _et al._ [-@ehigiamusoe2022nexus] found that while higher income inequality lowers carbon emissions and ecological footprints in high-income countries, it worsens them in middle-income states and the effect of poverty reduction on environmental outcomes seems to dominate in lower-income countries. Similarly, Grunewald [-@Grunewald2017] found that income inequality and carbon emissions were negatively correlated in low- and middle-income countries but positively correlated in high-income countries.

Our findings tend to support the conclusion that lowering economic inequality and lowering environmental impact are synergistic. In particular, in our three-dimensional analyses of country trajectories within the trilemma space, no country started or ended in a region with (relatively) high inequality and high ecological footprint (the upper right quadrant of the cube). And based on their trajectories, all countries first decreased in income inequality before increasing in ecological footprint, regardless of prosperity level or magnitude of displacement from their starting point. This suggests that countries still have developmental "room" to decrease income inequality without increasing environmental impact (Figure 1G; Figure 2C).

Nonetheless, we recognize that the divergent trajectories (including both direction and magnitude of displacement) of countries and pronounced clustering, as well as the large variations in indicator performance in the static analyses, highlight the importance of structural international differences. This applies not only to geographic factors (e.g., population density, natural resource endowments, and climatic conditions determine per-capita biocapacity) but also to past development trajectories and the position of different countries in the structure of the world economy. Figge _et al._ [-@figge2017effects] found that even after controlling for prosperity (average income), globalization significantly increased countries’ ecological footprints globally. And based on analysis of per-capita ecological footprint and Gini for income, Teixidó-Figueras and Duro [-@teixido2015building] concluded that the primary approach to reduce overall global environmental impact is to reduce the large disparities between countries’ per-capita footprints through redistributive policies across countries, prioritizing international equity over economic growth per se. Therefore, structural reforms to international networks of trade and finance – which has come to be increasingly advocated by scholars, activists, policymakers in recent years – could be crucial to overcoming the sustainability trilemma.

Social-ecological feedbacks, which often reinforce the effects of geographic factors and development path dependencies, can also help explain the appearance of "regimes" such as the inequality trap found in the trilemma space. Weinzettel _et al._ [-@Weinzett:2013] found that countries with high per-capita biocapacity (i.e., large natural resource endowments relative to population level) tended to spare more land for nature, forming a virtuous cycle of improving environmental conditions. Similarly, population, environmental carrying capacity, and pollution have been found to have mutually-reinforcing feedbacks across countries [@LaTorr:2019]. This helps to explain the relatively good performance of countries such as Canada, Sweden, and Australia in our analyses. These types of social-ecological feedbacks can create and maintain "red loop" or "green loop" dynamics, entrapping some countries in long-term conditions of high poverty, inequality, and environmental degradation while placing others on a more sustainable trajectory [@Cumming:2014; @Cumming:2018].

@@

Whether the synergies and tradeoffs between prosperity, inequality, and the environmental impact identified in this study hold across scales (e.g., at subnational scales or in other analytic units such as cities or corporations), also remains an open question. Furthermore, the feedback mechanisms for interactions among the three dimensions at different scales are also likely to be complex [@Weinzett:2013; @Cumming:2018; @Hamann:2018; @LaTorr:2019]. If there is a unifying concern among countries and across scales, it may be the foundational role of nature, which provides the basic life-support systems for human well-being and the inescapable context for collective development. As our findings strongly suggest, a myopic focus on ever-growing prosperity as the solution to inequality and degradation is misplaced. Indeed, highly prosperous countries do not necessarily have lower levels of inequality, but almost always have higher environmental impacts. One undeniable trend identified in our analysis is that a singular focus on economic growth can undermine progress towards social equality and, most especially, environmental integrity [@arrow2004we; @fanning2022social].

@@

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{FigS4}
\caption{\textbf{Trajectories examples} As shown by the experiences of Ukraine and China from 1995 to 2017, historical and structural factors influence the trajectories of countries within the "trilemma space" formed by the economic, social, and environmental dimensions of sustainability: (A) the global financial crisis; (B) the beginning of the Russo-Ukrainian conflict; (C) the accession of China into the World Trade Organization; (D) the beginning of the large-scale anti-corruption crackdown in China.}
\label{fig:S4}
\end{figure*}
```

>>>>

- No country in the world simultaneously achieved high prosperity, low inequality, and low environmental impact between 1995-2017.
- There appears to be a global sustainability "trilemma" as opposed to a "triple bottom line".
- This global challenge was unevenly distributed, evidenced by countries’ divergent trajectories and clustering effects.
- There were widespread tradeoffs among the three indicators, with the prosperity-environment tradeoff most pronounced.
- Mitigating the prosperity-environment tradeoff is made more difficult by high inequality.

@@

A key aim of sustainable development is the joint achievement of prosperity, equality, and environmental integrity: in other words, material living standards that are high, broadly-distributed, and low-impact. This has often been called the "triple bottom line". But instead, what if there is a "trilemma" that inhibits the simultaneous achievement of these three goals? We analyzed international patterns and trends in the relationships between per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint from 1995 to 2017, benchmarking them against thresholds from the existing literature. A "dynamic" analysis of the trajectories of 59 countries and a "static" analysis of a larger sample of 140 countries found that none met the triple bottom line, and that instead there were widespread tradeoffs among the three indicators. These tradeoffs, leading to divergent national trajectories and country clusters, show that common pair-wise explanations such as Kuznets Curves do not adequately capture important development dynamics. In particular, while only a few countries simultaneously met the thresholds for prosperity and equality on the one hand and equality and environment on the other, none did for prosperity and environment. Moreover, inequality likely makes resolving this critical tradeoff more difficult. Our findings suggest that mitigating the sustainability trilemma may require countries - especially those that are already prosperous - to prioritise economic redistribution and environmental stewardship over further growth.

@@

Although global prosperity has increased in recent generations, progress has been uneven. Even today, a significant share of the world’s population lives under the international poverty line (a standard that the UN itself recognizes may grossly undercount the world's impoverished [@Alston_20220]). Right above this derisory threshold, hundreds of millions more people maintain a precarious subsistence and aspire for higher and more secure living standards [@TheWor:2018]. And while average incomes for certain developing countries have been converging with those of advanced economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened across the world [@Bourguignon:2007; @Milanovic:2016]. This has made  inequality an urgent issue for policymakers and the general public due to  anxieties about social and political instability [@Chancel:2020; @UN2020]. Environmental integrity forms the third, and perhaps most fundamental, dimension of the sustainability [@arrow1995economic]. There is now a scientific consensus that human activities have had severely deleterious impacts on the biosphere, the health of which determines long-run human wellbeing [@Diaz:2019; @folke2021our]. Indeed, humanity has already exceeded a number of the biophysical boundaries that delimit a "safe operating space" for humanity [@Rockström:2009; @RaworthKate:2017; @Rockstr_m_2023].

@@

We used these data to chart the movement of countries across the dimensions of prosperity, inequality, and environmental impact over time. We further investigated whether these movements trace out typologies of country trajectories within the "trilemma space". To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time. We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to determine the relevant number of clusters (clustering is a statistical process whereby objects are partitioned and then grouped in such a way that the members of one cluster are more closely related to each other than they are to those of other clusters). We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by Charrad _et al._, [-@Charrad:2014tp] and Brook _et al._ [-@Brock:2008vz]. We used principal component analysis as an ordination method for visualization, and then vector-fitting to show how clusters are pulled apart in the principal component space (in order to make the three-dimensional distribution of countries intelligible). All data used are publicly available, and the code used for the analysis is available at: <link will be added upon publication, avoided here due to double-blind requirements>.

While we were able to assess 140 countries "statically" (i.e., ignoring temporal dynamics), an analysis of the missing values showed that a "dynamic" analysis of changes over time in the trilemma space was only possible for only 59 countries (i.e., data for all three indicators were available for every year of the study period). The most incomplete dataset is that for inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values. Based on the quality of the three datasets, we selected the time-series from 1995-2017 for the 59 countries with fewer than 30\% of missing values and then imputed the missing values with linear interpolation. This reduced dataset enables the study of the temporal dynamics of countries in the trilemma space (Fig \ref{fig:trilemma}).

Finally, while global patterns provide insight into international variations in key dimensions of sustainability, they do not completely reveal the diversity of individual country trajectories, nor their typologies. To further investigate country trajectories, we also calculated their displacement within the trilemma space for each pair-wise relationship, where $\gamma$ (gamma) is the angle in the inequality and prosperity plane while $\theta$ (theta) is the angle in the inequality and environmental impact plane. These two angles and the total distance "travelled" in the three-dimensional space help to define the typologies of country trajectories.

@@

Our analyses found that no country simultaneously achieved high prosperity, low inequality, and low environmental impact. This strongly suggests that there is a sustainability trilemma at the global scale. More specifically, the dynamic analysis of 59 countries found that there were widespread tradeoffs between reducing inequality and reducing environmental impact (Fig \ref{fig:trilemma}E - as the Gini coefficient for income decreased, ecological footprint increased) as well as between increasing prosperity and reducing environmental impact (Fig \ref{fig:trilemma}G - per-capita GNI and ecological footprint moved in the same direction). The results also found persistent tradeoffs between prosperity and inequality (Fig \ref{fig:trilemma}G). Even several high-income countries (e.g., Sweden and New Zealand) that exhibited low levels of inequality in the past began to exhibit high variation and then a trend towards greater inequality with continued income growth.

@@

There are certain regions in the trilemma space where countries tended to dwell (Fig \ref{fig:cube_ensamble}). Multiple stable states can exist if feedbacks between development drivers are present. Statistically, this translates into multi-modal distributions and so-called "regimes". The prosperity-environment plane (Fig \ref{fig:cube_ensamble}A), and inequality-environment plane (Fig \ref{fig:cube_ensamble}C) have two pronounced clusters of countries (the movement of countries during the study period stayed within a relatively circumscribed space), while the inequality-prosperity plane (Fig \ref{fig:cube_ensamble}B) has one major cluster and one looser grouping. Countries also showed distinctive profiles in direction of movement (Fig \ref{fig:cube_ensamble} insets). Most countries had low displacement (short arrows), which means their trajectory over the study period did not take them far from where they started. A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995. For example, between 1995 and 2017, China's Gini coefficient for income and per-capita ecological footprint rose significantly alongside per-capita GNI. And while inequality and environmental impact fell in Ukraine, prosperity only increased marginally. To better understand country typologies, we clustered summary statistics (i.e., mean and standard deviations) of the three indicators, as well as the direction and magnitude of country movement over time (Fig \ref{fig:pca}).

The trajectories in the trilemma space of the 59 countries from 1995 to 2017 can be categorized into three general types (Fig \ref{fig:pca}A). First, there was a group composed of countries with relatively high prosperity and ecological footprint; their movements were relatively less erratic and had larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group). This cluster can be called "Regime A". Next, there was a group of countries, mainly from Latin America, that seemed to be "trapped" in a region of high inequality during the study period. They moved toward regions of higher income inequality with respect to the sample and with a larger standard deviation of $\theta$; that is to say, large variations along the income inequality and ecological footprint axes (Fig \ref{fig:pca}A, in green). Within the three-dimensional space, this cluster of countries with high inequality and relatively low prosperity and environmental impact can be called "Regime B". A third group of countries showed large displacements within the trilemma space but in no definite direction, and was characterized by large variations in prosperity and inequality (Fig \ref{fig:pca}A, in yellow).

If we ignore temporal dynamics and focus on the mean and standard deviation of the indicators over the study period, we are able to include more countries (N=140). This static analysis also yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between prosperity, inequality, and environmental impact as in the findings from the dynamic analysis. The first cluster (Fig \ref{fig:pca}B, in yellow) was composed primarily of advanced Western economies and had high prosperity, inequality, and environmental impact. The second cluster (Fig \ref{fig:pca}B, in green) was composed of countries from Africa, Latin America, and Asia - broadly, the "Global South". Here, countries all tended to have lower prosperity (e.g., low- or middle-income status), but among them exhibited large variations in inequality and environmental impact.

In both the dynamic and the static analysis, decreasing inequality and decreasing environmental impact appear to pull countries in opposite directions (Fig \ref{fig:pca}), providing evidence for the difficulty of simultaneously achieving both goals. In Fig \ref{fig:pca}A, the fitted vectors (arrows) show the direction in which variables explain the variance in the data and thus countries' differences. While per-capita ecological footprint and GNI point in similar directions (countries tended to increase both at the same time), the Gini for income points in the opposite direction. This shows that a few  countries that reduce inequality did so only after increasing in environmental impact and prosperity. In the ordination performed in the larger static sample (Fig \ref{fig:pca}B), we find a similar pattern. In fact, the trajectories analysis found that countries that reduced inequality subsequently increased in environmental impact (Fig \ref{fig:cube_ensamble}A). Overall, the findings of both the dynamic and the static analysis show that there are tradeoffs between the three "bottom lines" of prosperity, equality, and environmental integrity at the global level. This is perhaps most simply illustrated by the fact that in both analyses, only a small subset of countries simultaneously met the prosperity and inequality thresholds on the one hand and the inequality and environment thresholds on the other, and no countries simultaneously met the prosperity and environment thresholds (Fig \ref{fig:trilemma} E-G; Fig \ref{fig:cube_ensamble} A-C).

@@

Various arguments have been advanced to explain the relationships between prosperity, equality, and environmental integrity in global development. Most commonly, mechanisms have been hypothesized for pair-wise relationships, with Kuznets Curve-type relationships being the most widely analyzed. The (Environmental) Kuznets Curve posits that growing prosperity leads to worsening inequality and environmental impact, before reaching an inflection point after which the is reversed. The evidence for this hypothesis has been contested with respect to both inequality [@Chancel:2020; @Pikett:2014] and the environment [@stern2004rise; @stern1996economic ]. In a survey of the empirical literature, Chancel [-@Chancel:2020] concluded that evidence supporting (Environmental) Kuznets Curve-type relationships tended to be statistical artefacts particular to post-Second World War datasets or transient historical trends. Therefore, such findings have limited applicability to today’s conditions of large-scale social-ecological change.

Our findings support these arguments: neither the dynamic nor the static analysis detected Kuznets Curve relationships between prosperity and either the Gini coefficient for income or ecological footprint Fig \ref{fig:S5}) (please see SI for more details). Instead, the findings strongly suggest that growing prosperity is almost always associated with worsening inequality and environmental impact. The low frequency with which countries were able to meet even two goals in both the dynamic and the static analysis is highly notable. Indeed, the fact that no country simultaneously met the prosperity and environment thresholds suggests that resolving the "dilemma" of this particular tradeoff is the necessary (but insufficient) condition for resolving the broader sustainability trilemma facing global development.

A growing body of empirical studies have sought to identify potential "decoupling" between economic growth and various indicators of environmental impact. The evidence for such trends have been mixed. In a review of the literature going back thirty years, Vadén _et al_. [-@vaden2020decoupling] found evidence of global decoupling between economic growth and certain pollutants, as well as country-level cases of decoupling from land and water use. But evidence of economy-wide decoupling from resource use at the international scale has proven elusive. However, Szigeti et al. [-@szigeti2017decoupling] did identify a positive international trend in the decreasing ecological footprint intensity of economic growth over the past two decades. But whether this trend can continue to the point where the prosperity-environment tradeoff is fully mitigated remains an open and pressing question. Our findings suggest that this dilemma could remain a major hurdle for global development into the foreseeable future. Breaking the link will likely require large-scale changes in governance and cultural norms, including replacing the policy primacy of economic growth with the prioritization of social and environmental goals, especially in already affluent communities and countries [@wiedmann2020scientists].

Our findings also show that inequality is likely to make mitigating prosperity-environment tradeoffs more difficult, as it appears to "pull" country trajectories away from the prosperity and environment thresholds in the trilemma space (Fig \ref{fig:trilemma}E-G). With respect to the prosperity dimension, there is strong empirical evidence showing that economic inequality tends to slow growth for a variety of institutional and behavioural reasons, from rent-seeking and the political influence of vested interests to the basic microeconomic logic that the marginal propensity to consume is inversely related to affluence. In other words, a more unequal economy is likely to have relatively lower aggregate demand [@stiglitz2016inequality; @carvalho2016personal; @aguiar2015has]. The widespread tradeoffs between per-capita GNI and the Gini coefficient for income discovered in our analyses support this conclusion.

The relationship between inequality and the environment is more complicated. Two primary types of mechanisms have been posited in the literature, one "economic" and the other "political" [@berthe2015mechanisms; @Hamann:2018]. Similar to inequality’s effect on economic growth, the former hypothesizes that higher inequality lessens environmental impact because it reduces overall consumption (this has sometimes been called the "marginal propensity to emit" effect). The political hypothesis argues that, by concentrating power in the hands of the rich, inequality worsens environmental conditions because the well-heeled tend to have a vested interest in stopping profit-lowering regulations that protect the environment [@boyce1994inequality; @KEMPF_2007]. The tradeoffs between income inequality and environmental impact in our findings support the economic hypothesis.

In the existing literature, these tradeoffs have been shown for countries across the geographic and development spectrum. In a study of Western African states from 1984 to 2016, Langnel _et al._ [-@langnel2021income] found that income inequality often lowers ecological footprint, and Ali [-@ali2023income] discovered a tradeoff between the Gini for income and carbon emissions in 42 middle-income countries between 1990 and 2016. In high-income economies, Hailemariam _et al._ [-@hailemariam2020carbon]’s study of OECD countries found that increasing income inequality (also measured by the Gini coefficient) was associated with decreasing carbon emissions. Similar to our findings, but only focused on a subset of Sub-Saharan countries from 1995 to 2018, Gimba _et al._ [-@gimba2023towards] found that rising income inequality reduces ecological footprint but exerts a significant drag on economic growth – with critical implications for poverty alleviation in a region with many of the world’s lowest-income countries.

Our analyses showed that the development burden of the sustainability trilemma was distributed highly unevenly across countries. For instance, although no country simultaneously met the triple bottom line thresholds, some improved along two dimensions and a few – notably Belgium and New Zealand – advanced in all three. This heterogeneity of international performance is also underscored by the existence of country clusters or regimes (Fig \ref{fig:cube_ensamble}). Most notably, a group of countries primarily from Latin America struggled to escape an area in the trilemma space characterised by high inequality ("Regime 2"); this seems to have also hindered their progress in the other two dimensions, further supporting the idea that inequality makes mitigating the prosperity-environment tradeoff more difficult.

Two plausible sets of factors may explain this unevenly-distributed burden of the sustainability trilemma: geographic heterogeneity and past development trajectories. Weinzettel_et al. [-@Weinzett:2013] found that countries with high per-capita biocapacity (i.e., large natural resource endowments relative to population level) tended to spare more land for nature, forming a virtuous cycle of improving ecological conditions. Similarly, population, environmental carrying capacity, and pollution have been found to exert mutually-reinforcing feedbacks internationally [@LaTorr:2019]. These types of social-ecological feedbacks can create "red loop" or "green loop" dynamics, entrapping some countries in conditions of high poverty, inequality, and environmental degradation while placing others on a more sustainable trajectory [@Cumming:2014; @Cumming:2018]. In our analyses, these mechanisms may help explain the relatively good performance of certain affluent economies such as Canada, Sweden, and Australia on the one hand and the inequality-induced clustering of Latin American countries on the other.

Finally, the relative positions of countries in the structure of the world economy can also shed light on divergent international trajectories with respect to the three indicators. In an earlier study, Figge _et al._ [-@figge2017effects] found that even after controlling for average income, globalization significantly increased countries’ ecological footprints. And based on analysis of per-capita ecological footprint and the Gini coefficient for income, Teixidó-Figueras and Duro [-@teixido2015building] concluded that reducing global environmental impact requires reducing the large disparities between countries, including through redistributive policies that prioritize international equity over the growth of individual economies. Therefore, in addition to domestic policies that place greater emphasis on equality and environmental integrity, structural reforms to international networks of trade and finance may also be needed to mitigate the sustainability trilemma.

@@

@@
